---
title: "Hypothesis 2 & 1"
highlighter: prettify
output: 
  html_document:
    code_folding: hide
    toc: yes
    toc_depth: 2
    toc_float: yes
    theme: readable
    includes:
      after_body: footer.html   
---
<style>
body{
  font-family: 'Oxygen', sans-serif;
  font-size: 16px;
  line-height: 24px;
}
</style>


```{r, echo=FALSE, eval=F}
htmltools::img(src = knitr::image_uri(file.path("C:/Users/k1894405/OneDrive - King's College London/PhD/Projects/Genomic SEM project/", "KCL_logo.jpg")), 
               alt = 'logo', 
               style = 'position:absolute; top:0; right:110px; padding:20px;')
```

```{r setup, include=F,echo=F}
require(knitr)
opts_knit$set(root.dir = "C:/Users/k1894405/OneDrive - King's College London/PhD/Projects/Genomic SEM project/")
knitr::opts_chunk$set(echo = TRUE, fig.align="center")
```

$$\\[0.5in]$$

Here we compare genetic and phenotypic brain networks and investigate whether they have the same properties in terms of their association with age-volume correlations. 


***
<br>
<br>

# Hypothesis 2
We hypothesise a positive association and high congruence between the two sets of
factor loadings which would implicate that the same dimensions that underlie
phenotypic variation in regional volumes also control the dimension of genetic sharing
across these brain volumes. That is, it would suggest that phenotypic structural
integrity recapitulates genetic integrity and that they are underpinned by similar
general biological mechanisms.
## Association between phenotypic and genetic correlations

Previously, we have calculated a phenotypic and genetic correlation matrix between 83 brain areas. Here we test whether phenotypic correlations were equivalent to genetic correlations.
```{r pheno_vs_geno,warning=F,message=F}
##### correlate genetic and phenotypic correlations between brain regions
##### load genetic data
#set wd
workingd<-getwd()
temporarywd<-paste0(workingd,"/data_my_own/ldsc/")
setwd(temporarywd)

load("whole_brain.RData")
ldscoutput<-LDSCoutput_wholebrain

# name rows and columns
dimnames(ldscoutput$S_Stand)[[1]]<-dimnames(ldscoutput$S)[[2]]
dimnames(ldscoutput$S_Stand)[[2]]<-dimnames(ldscoutput$S)[[2]]

# keep lower matrix
cormatrix<-ldscoutput$S_Stand
get_lower_tri<-function(cormatrix){
  cormatrix[upper.tri(cormatrix)] <- NA
  return(cormatrix)
}
genetic_cor<-get_lower_tri(cormatrix)
genetic_cor<-reshape2::melt(genetic_cor)
names(genetic_cor)<-c("region1","region2","genetic_cor")

######## load phenotypic data

temporarywd_pheno<-paste0(workingd,"/data_my_own/Pheno_preparation/")
setwd(temporarywd_pheno)

load("pheno_decomposition.RData")

phenotypic_cor<-cor_matrix

# sort same as genetic matrix
names<-dimnames(ldscoutput$S)[[2]]
phenotypic_cor<-phenotypic_cor[names,names]

# get lower triangle
phenotypic_cor<-get_lower_tri(phenotypic_cor)
phenotypic_cor<-reshape2::melt(phenotypic_cor)
names(phenotypic_cor)<-c("region1","region2","phenotypic_cor")

gen_pheno<-merge(genetic_cor,phenotypic_cor,by=c("region1","region2"))



# remove correlations with itself (genetic and phenotypic)
gen_pheno$genetic_cor<-ifelse(gen_pheno$region1 == gen_pheno$region2,NA,gen_pheno$genetic_cor)
gen_pheno$phenotypic_cor<-ifelse(gen_pheno$region1 == gen_pheno$region2,NA,gen_pheno$phenotypic_cor)
# calculate correlation between phenotypic and genetic correlations
#cor.test(gen_pheno$genetic_cor,gen_pheno$phenotypic_cor)
cor_cor<-round(cor(gen_pheno$genetic_cor,gen_pheno$phenotypic_cor,use="pairwise.complete.obs"),digits=2)
#plot(gen_pheno$genetic_cor,gen_pheno$phenotypic_cor,ylab="Phenotypic correlations",xlab="Genetic correlations",xlim=c(-0.3,1),ylim=c(-0.3,1))
#abline(lm(gen_pheno$phenotypic_cor~gen_pheno$genetic_cor,na.action = "na.omit"),col="blue")
#abline(0,1,col="red")
#abline(x=0,col="green")

library(ggplot2)
library(ggpubr)

plot_corr<-ggplot(gen_pheno, aes(x=genetic_cor,y=phenotypic_cor))+#
  geom_point(size=4,alpha=0.4)+
  geom_smooth(method=lm, se=TRUE) #+
  #ggtitle("Phenotypic vs. genetic correlations")
  
plot_corr<- plot_corr+xlab("Genetic correlations")+ ylab("Phenotypic correlations")+
  theme(plot.title = element_text(size=20, hjust = 0.5), panel.background = element_blank(), axis.line = element_line(color="black"), axis.line.x = element_line(color="black"))+ theme_bw()+
    stat_cor(method="pearson",cor.coef.name="r",size=4.5,position = "identity",)+
  theme(legend.position = "none")+
    geom_abline(intercept=0,slope=(1),linetype="dashed",color="red",show.legend=T, size=1,alpha=0.7)+
    geom_vline(xintercept = 0,linetype="dashed",color="black",size=1.5,alpha=0.2)+
    theme(text = element_text(size=12,family="Arial Narrow"),
        axis.text.x = element_text(size=15,family="Arial Narrow"),#angle=45
        axis.text.y = element_text(size=15,family="Arial Narrow"),
        axis.title.y = element_text(face="bold", colour='black', size=17,family="Arial Narrow"),
        axis.title.x = element_text(face="bold", colour='black', size=17,family="Arial Narrow"))
plot_corr

```

**The red line is the line of identity.**


We correlated phenotypic with genetic correlations among these areas, and we obtained a significant correlation of `r cor_cor`. This indicates that phenotypic correlations predict genetic correlations fairly well.

<br/>
<br/>


```{r neg_SEs}
temporarywd<-paste0(workingd,"/data_my_own/ldsc/")
setwd(temporarywd)

load("whole_brain.RData")
ldscoutput<-LDSCoutput_wholebrain
  dimnames(ldscoutput$S_Stand)[[1]]<-dimnames(ldscoutput$S)[[2]]
  dimnames(ldscoutput$S_Stand)[[2]]<-dimnames(ldscoutput$S)[[2]]

k<-nrow(ldscoutput$S)
SE<-matrix(0, k, k)
SE[lower.tri(SE,diag=TRUE)] <-sqrt(diag(ldscoutput$V))

get_lower_tri<-function(cormatrix){
  cormatrix[upper.tri(cormatrix)] <- NA
  return(cormatrix)
}

SE_lower<-get_lower_tri(SE)
SE_lower<-reshape2::melt(SE_lower)
```



***
<br>

## Association between phenotypic and genetic loadings on the first PC
```{r pheno_gen_stand_loadings, warning=F}
# load phenotypic loadings
temporarywd_pheno<-paste0(workingd,"/data_my_own/Pheno_preparation/")
setwd(temporarywd_pheno)

load("pheno_decomposition.RData")

pheno_stand_loadings<-stand_loadings
names(pheno_stand_loadings)[2]<-"pheno_stand_loadings"
pheno_stand_loadings$pheno_stand_loadings<-pheno_stand_loadings$pheno_stand_loadings*(-1)

# load genetic loadings
library(data.table)

temporarywd<-paste0(workingd,"/data_my_own/standardised_loadings/")
setwd(temporarywd)

gen_stand_loadings<-read.table("stand_loadings_whole_brain.txt",header=T)
names(gen_stand_loadings)[2]<-"gen_stand_loadings"

	# multiply the loadings with -1 if the median of the loadings is negative
	median_loadings<-median(gen_stand_loadings$gen_stand_loadings)
	mean_direction<-sign(median_loadings)
	if(mean_direction == -1){
	gen_stand_loadings$gen_stand_loadings<-gen_stand_loadings$gen_stand_loadings*(-1)
	} else {
	gen_stand_loadings$gen_stand_loadings<-gen_stand_loadings$gen_stand_loadings}
	
	
stand_loadings<-merge(pheno_stand_loadings,gen_stand_loadings,by="Regions")

#calculate correlation between loadings
summary(lm(stand_loadings$pheno_stand_loadings~stand_loadings$gen_stand_loadings,na.action = "na.omit"))
# save statistics in objects for text
model<-lm(stand_loadings$pheno_stand_loadings~stand_loadings$gen_stand_loadings,na.action = "na.omit")
beta<-round(summary(model)$coefficient[2,1],digits = 2)
SE<-round(summary(model)$coefficient[2,2],digits = 2)
pval<-signif(summary(model)$coefficient[2,4],digits =3)


## calculate absolute mean difference between phenotypic and genetic PC loadings
#delta_cor<-abs(stand_loadings$pheno_stand_loadings-stand_loadings$gen_stand_loadings)
#mean(delta_cor,na.rm=T)
#sd(delta_cor,na.rm=T)
```
Genetic loadings significantly predict phenotypic loadings (*b* = `r beta`, *SE* = `r SE`, *pvalue* = `r pval`).


```{r plot_loadings, warning=F,message=F,results="hide",figure.keep="all"}
library(ggplot2)
library(ggpubr)

plot_loadings<-ggplot(stand_loadings, aes(x=gen_stand_loadings,y=pheno_stand_loadings))+#
                        geom_point(size=4,alpha=0.4)+#,color=stand_loadings$central_col
                        geom_smooth(method=lm, se=TRUE) #+
                        #ggtitle("Phenotypic vs. genetic PC loadings")
  
  plot_loadings<-plot_loadings+
                xlab("Genetic PC loadings")+
                ylab("Phenotypic PC loadings")+
                theme(
                  plot.title = element_text(size=20, hjust = 0.5), 
                  panel.background = element_blank(), 
                  axis.line = element_line(color="black"), 
                  axis.line.x = element_line(color="black"))+ 
    theme_bw()+
    stat_cor(method="pearson",cor.coef.name="r",size=4.5,label.x.npc=0.65,label.y.npc = 0)+
    theme(legend.position = "none")+
    geom_abline(intercept=0,
                slope=(1),
                linetype="dashed",
                color="red",
                show.legend=T, 
                size=1.5,
                alpha=0.6)+
        theme(text = element_text(size=12),
              axis.text.x = element_text(size=15),
              axis.text.y = element_text(size=15),
              axis.title.y = element_text(face="bold", colour='black', size=17),
              axis.title.x = element_text(face="bold", colour='black', size=17))
plot_loadings
```

**The red line is the line of identity.**

***
<br>

## Tucker congruence coefficient
### Tucker coefficient: Whole brain

```{r tucker_index,warning=F,message=F,results="hide"}
# to calculate factor congruence, insert covariance matrix
# cor_matrix saved in pheno_decomposition.RData is phenotypic cov matrix
# https://openpsych.net/forum/attachment.php?aid=364
# cut off for Tucker index to assume similarity range from .85-.94
library(psych)
pheno_eigenvector<-principal(cor_matrix)
match_names<-dimnames(cor_matrix)[[1]]

# genetic matrix inferred through LDSC 
temporarywd<-paste0(workingd,"/data_my_own/ldsc/")
setwd(temporarywd)

load("whole_brain.RData")
genetic_whole_brain_matrix<-LDSCoutput_wholebrain$S_Stand
colnames(genetic_whole_brain_matrix)<-dimnames(LDSCoutput_wholebrain$S)[[2]]
rownames(genetic_whole_brain_matrix)<-dimnames(LDSCoutput_wholebrain$S)[[2]]
genetic_whole_brain_matrix<-genetic_whole_brain_matrix[match_names,match_names]
gen_eigenvector<-principal(genetic_whole_brain_matrix)


tucker_index<-factor.congruence(x=pheno_eigenvector,y=gen_eigenvector)
# smoothing required because the last PC has negative eigenvalues
```

The Tucker congruence coefficient for the loadings of the first PC in phenotypic vs. genetic correlation matrices is **`r tucker_index`**. It has been previously suggested that an index between .85-.94 indicates a "fair similarity" between the loadings (Lorenzo-Seva & ten Berge, 2006, *Methodology*, doi: 10.1027/1614-1881.2.2.57).


### Tucker coefficient: A priori networks
```{r Tucker_networks, warning=F,message=F}
# read in genetic corr matrix
temporarywd<-paste0(workingd,"/data_my_own/ldsc/")
setwd(temporarywd)

load("whole_brain.RData")
genetic_whole_brain_matrix<-LDSCoutput_wholebrain$S_Stand
colnames(genetic_whole_brain_matrix)<-dimnames(LDSCoutput_wholebrain$S)[[2]]
rownames(genetic_whole_brain_matrix)<-dimnames(LDSCoutput_wholebrain$S)[[2]]

# set wd to where region names per network are saved
temporarywd<-paste0(workingd,"/Scripts/my_own_gwas/2Perform_ldsc/")
setwd(temporarywd)

# load files containing region names
central_exec_regions<-read.table("central_executive_areas.txt")$V1
cingulo_regions<-read.table("cingulo_opercular_areas.txt")$V1
default_regions<-read.table("default_mode_areas.txt")$V1
hippocampal_regions<-read.table("hippocampal_diencephalic_areas.txt")$V1
multiple_demand_regions<-read.table("multiple_demand_areas.txt")$V1
p_fit_regions<-read.table("p_fit_areas.txt")$V1
salience_regions<-read.table("salience_areas.txt")$V1
sensori_regions<-read.table("sensorimotor_areas.txt")$V1
temporo_regions<-read.table("temporo_amygdala_orbitofrontal_areas.txt")$V1


library(psych)
library(stringr)
network_regions<-ls(pattern="_regions")

# create df to display Tucker-Lewis index per network
display_Tucker<-data.frame(matrix(ncol=2))
colnames(display_Tucker)<-c("network","Tucker_index")

for(i in network_regions){
  regions<-get(i)
  #print(i)
  
  # decomposition pheno
  pheno<-principal(cor_matrix[regions,regions])

  # decomposition geno
  geno<-principal(genetic_whole_brain_matrix[regions,regions])
  
  # calculate Tucker-Lewis index
  tucker_lewis_index<-factor.congruence(x=pheno,y=geno)
  
  # save index
  network_name<-str_remove(i,pattern="_regions")
  #assign(network_name,tucker_lewis_index)
  
  # save index in table
  display_Tucker[which(network_regions ==i),"network"]<-network_name
  display_Tucker[which(network_regions ==i),"Tucker_index"]<-tucker_lewis_index
}

kable(display_Tucker, label="Tucker-Lewis Index indicating factor congruence between phenotypic and genetic networks", col.names = c("Networks","Tucker congruence coefficient"))
```

***

<br>

# Hypothesis 1

We hypothesise a negative correlation between genetic brain-wide PC loadings and
phenotypically derived age sensitivities. We do not make a prediction about the size of
the correlation as no previous study known to us has investigated this question on a
genetic level. A significant association would indicate that the same general
dimensions of variation in global brain aging underlie the dimension of genetic
sharing across brain regions.

## Association between phenotypic PC loadings and age-volume correlations


```{r corr_phenotypes, message=F,warning=F}
# load phenotypic loadings
names(age_cor)[2]<-"Regions"
phenotest<-merge(stand_loadings,age_cor,by="Regions")

plot_age_pheno<-ggplot(phenotest, aes(x=pheno_stand_loadings,y=age_cor))+#
  geom_point(size=4,alpha=0.4)+
  geom_smooth(method=lm, se=TRUE)#,color="blue") #+
  #ggtitle("Phenotypic age correlations vs. phenotypic PC loadings")
  
  plot_age_pheno<-
    plot_age_pheno+
    xlab("Phenotypic PC loadings")+
    ylab("Age correlations")+
    theme(plot.title = element_text(size=20, hjust = 0.5), 
          panel.background = element_blank(), 
          axis.line = element_line(color="black"), 
          axis.line.x = element_line(color="black"))+ 
    theme_bw()+
    stat_cor(method="pearson",
             cor.coef.name="r",
             size=4.5,
             color="black",
             label.x.npc=0.65,
             label.y.npc = 0)+
  theme(legend.position = "none")+
        theme(text = element_text(size=12),
        axis.text.x = element_text(size=15),#angle=45
        axis.text.y = element_text(size=15),
        axis.title.y = element_text(face="bold", colour='black', size=17),
        axis.title.x = element_text(face="bold", colour='black', size=17))

plot_age_pheno
```



## Association between genetic PC loadings and age-volume correlations

```{r age_corr}
# merge age  correlations with standardised loadings
names(age_cor)[2]<-"Regions"
data<-merge(stand_loadings,age_cor,by="Regions")

# correlate genetic stand loadings with age cor
age_correlation_result<-cor.test(data$gen_stand_loadings,data$age_cor)
age_cor_estimate<-round(age_correlation_result$estimate,digits = 2)
age_cor_pval<-round(age_correlation_result$p.value,digits = 3)

```

Genetic loadings do not significantly correlate with age sensitivity (*r* = `r age_cor_estimate`, *p-value* = `r age_cor_pval`).

```{r lot_age_cor,echo=F,message=F,warning=F}
plot_age_geno<-ggplot(data, aes(x=gen_stand_loadings,y=age_cor))+#
  geom_point(size=4,alpha=0.4)+#,color=data$central_col
  geom_smooth(method=lm, se=TRUE) #+
  #ggtitle("Phenotypic age correlations vs. PC genetic loadings")
  
plot_age_geno<-
  plot_age_geno+
  xlab("Genetic PC loadings")+
  ylab("Age correlations")+
  theme(plot.title = element_text(size=20, hjust = 0.5), 
        panel.background = element_blank(), 
        axis.line = element_line(color="black"), 
        axis.line.x = element_line(color="black"))+ 
  theme_bw()+
    stat_cor(method="pearson",
             cor.coef.name="r",
             size=4.5,
             label.x.npc=0.65,
             label.y.npc = 0)+
      theme(text = element_text(size=12),
        axis.text.x = element_text(size=15),#angle=45
        axis.text.y = element_text(size=15),
        axis.title.y = element_text(face="bold", colour='black', size=17),
        axis.title.x = element_text(face="bold", colour='black', size=17))
plot_age_geno

```

### Plot all four correlation plots
```{r merge_plots_gen_pheno,eval=F,message=F,warning=F}
library(cowplot)

tiff("corr.tiff", width = 15, height = 10, units = 'in', res=1000)

plot_grid(plot_corr,plot_loadings,plot_age_pheno,plot_age_geno,nrow = 2,ncol = 2,labels=c("A","B","C","D"),width=1,height=1, label_size = 18,hjust = 0.07,scale = 0.85)

dev.off()
```


