---
title: "Processing genetic data"
highlighter: prettify
output: 
  html_document:
    code_folding: hide
    toc: yes
    toc_depth: 2
    toc_float: yes
    theme: readable
    includes:
      after_body: footer.html   
---
<style>
body{
  font-family: 'Oxygen', sans-serif;
  font-size: 16px;
  line-height: 24px;
}
</style>


```{r setup, include=F,echo=F}
require(knitr)
opts_knit$set(root.dir = "C:/Users/k1894405/OneDrive - King's College London/PhD/Projects/Genomic SEM project/")
knitr::opts_chunk$set(echo = TRUE, fig.align="center")
```

$$\\[0.5in]$$

*Here we analyse that GWAS summary statistics created as part of this project. The code displayed below performs data visualisation and sanity checks.*

$$\\[1in]$$

***



# 1. Munge files as expected by Genomic SEM software
```{r munge_gwas,eval=F,echo=T}
### this script will munge the gwas files using GenomicSEM

# load dependencies
library(tidyr)
library(data.table)
library(devtools)
library(GenomicSEM)
library(stringr)

# set wd to where formatted files are saved
setwd("/mnt/lustre/groups/ukbiobank/Edinburgh_Data/usr/anna/PhD/output/regenie/step2/")

hm3<-"/scratch/users/k1894405/genetic_networks/data/w_hm3.noMHC.snplist"


# list gwas files you want to munge
gwas_files<-list.files(pattern="GWAS_22chr_noTBVcontrol_")
print(length(gwas_files))

						

for (i in gwas_files){
		# define file to be munged
			files<-i
			print(files)
		
		# name the file
			trait_name <- str_remove(i, pattern = "GWAS_22chr_noTBVcontrol_")
			trait_name <- str_remove(trait_name, pattern = ".txt")

			trait.names<-trait_name
			print(trait.names)
		
		# munge
			munge(files=files,
					hm3=hm3,
					trait.names=trait.names,
					info.filter = 0.9)
}

```
After munging the files have 9 columns, and 1,171,881 rows (SNPs).

***

# 2. Calculate LDSC
```{r calculate_ldsc_83_regions,eval=F,echo=T}
### after formatting and munging the gwas files, I have manually transferred the files into a folder called "GWAS_munged"

library(stringr)
library(devtools)
library(GenomicSEM)

# set wd to the munged GWAS folder
setwd("/mnt/lustre/groups/ukbiobank/Edinburgh_Data/usr/anna/PhD/output/genetic_networks_project/gwas_munged/")

# specifications for ldsc function
traits<-list.files(pattern=".sumstats.gz")

# double check that 83 files have been read in 
if(length(traits) !=83){print("You are not including 83 files");break}

# no sample or population prevalence needed because brain volumes are continuous traits
# specify NA for both
desired_length<-length(traits)
sample.prev<-rep(NA, desired_length)
print(sample.prev)
population.prev<-rep(NA, desired_length)
print(population.prev)

# ld scores and weights previously downloaded (European population)
ld<-"/scratch/users/k1894405/genetic_networks/data/eur_w_ld_chr/"
wld<-"/scratch/users/k1894405/genetic_networks/data/eur_w_ld_chr/"


gwas_munged<-list.files(pattern=".sumstats.gz")
gwas_munged<-str_replace(gwas_munged,pattern=".sumstats.gz",replacement="")
trait.names<-gwas_munged

# double check that trait.names and traits match up
# first build a data set from the vectors, then compare the two vectors
test<-data.frame(trait.names,traits)
print(test)


count<-0

for(i in 1:nrow(test)){
	# which trait are we testing
		print(substr(test$traits[i], 1, nchar(as.character(test$trait.names[i]))))
		print(as.character(test$trait.names[i]))
	# break the loop if trait.names and traits don't match
		if(as.character(test$trait.names[i]) != substr(test$traits[i], 1, nchar(as.character(test$trait.names[i])))){"Trait names and trait files don't match";break}
		
		count<-count+1
		print(paste0("This is the ",count,"st/nd trait iteration"))
	}


LDSCoutput_wholebrain<-ldsc(traits=traits,
                             ld=ld,wld=wld,
                             trait.names = trait.names,
							 ldsc.log="/mnt/lustre/groups/ukbiobank/Edinburgh_Data/usr/anna/PhD/output/genetic_networks_project/ldsc/ldsc_wholebrain.log",
							 sample.prev=sample.prev,
							 population.prev=population.prev,
							 stand=T)

## save output for subsequent analyses
save(LDSCoutput_wholebrain, file="/mnt/lustre/groups/ukbiobank/Edinburgh_Data/usr/anna/PhD/output/genetic_networks_project/ldsc/ldsc_wholebrain.RData")
```

***


# 3. Plot heatmaps
```{r heatmaps, echo=T, out.width = '80%',warning=F,message=F}
## this script is displaying the heatmaps of the networks
rm(list=ls())
# load dependencies
library(stringr)
library(reshape2)
library(ggplot2)

# load data and name it according to network
workingwd<-getwd()
temporarywd<-paste0(workingwd,"/data_my_own/ldsc/")
setwd(temporarywd)

networks<-list.files(pattern=".RData")
network_names<-str_replace(networks,pattern=".RData",replacement = "")

for(i in 1:length(networks)){
load(networks[i])
  name<-network_names[i]
  assign(name,LDSCoutput)
}

## name all columns in S_Stand after brain regions and round S_Stand
for(i in network_names){
  output<-get(i)
  dimnames(output$S_Stand)[[1]]<-dimnames(output$S)[[2]]
  dimnames(output$S_Stand)[[2]]<-dimnames(output$S)[[2]]
  name<-i
  assign(name,output)
  output$S_Stand<-round(output$S_Stand,digits = 2)
  name_cor<-paste0("cor_",i)
  assign(name_cor,output$S_Stand)
}

# count number of correlation matrices
#length(ls(pattern="cor_"))

# create vector containing names of the correlation matrices
matrices<-ls(pattern="cor_")

## plot the rounded correlation matrices

# get lower triangle of matrix
for(i in matrices){
  cormatrix<- get(i)
  get_lower_tri<-function(cormatrix){
    cormatrix[upper.tri(cormatrix)] <- NA
    return(cormatrix)
  }
  lower_triangle<-get_lower_tri(cormatrix)
  lower_triangle<-reshape2::melt(lower_triangle)
  #print(i)
  #print(lower_triangle)
  lower_triangle$value<-ifelse(lower_triangle$value >1,1,lower_triangle$value)
  
  heatmap<-
    ggplot(data=lower_triangle, aes(Var1,Var2,fill=value))+
    geom_tile(color = "white")+
    theme_minimal()+ 
    theme_bw()+
    theme(axis.text.x = element_text(angle = 45, vjust = 1, 
                                     size = 8, hjust = 1))+
    theme(axis.text.y = element_text(vjust = 1, 
                                     size = 8, hjust = 1))+
    theme(axis.title.x=element_blank(),
          axis.title.y = element_blank(),
          panel.grid.major = element_blank(),
          panel.border = element_blank(),
          axis.ticks = element_blank(),
          legend.justification = c(1,0),
          #legend.position = c(0.45,0.7),
          legend.direction = "horizontal")+  
    scale_fill_gradient2(low="royalblue1",high="palevioletred",mid ="white",
                         midpoint=0,limit=c(-1,1),na.value="white",
                         name="Genetic\ncorrelation")+
    guides(fill=guide_colorbar(barwidth = 7,barheight = 1,title.position = "top",title.hjust = 0.5))+
    coord_fixed()
  
  heatmap<-heatmap+
    geom_text(aes(Var1,Var2,label=value),color="black",size=2)
  
  name<-paste0(i,"_heatmap")
  assign(name,heatmap)
}

```


### Example correlation matrices for three networks inferred through LDSC
```{r heatmaps1,echo=FALSE, out.width = '80%',warning=F,message=F,fig.cap=c("Central Executive","Cingulo-opercular")}
library(cowplot)
cor_central_exec_heatmap
cor_cingulo_heatmap
```

```{r heatmaps2,echo=FALSE, out.width = '100%',warning=F,message=F,fig.cap="Default Mode"}
cor_default_mode_heatmap
```

<br>
<br>

### Plot heatmap for genetic correlations across the whole brain

```{r whole_brain_heatmap,eval=T,echo=T,warning=F,message=F}
# set working directory
workingd<-getwd()
temporarywd<-paste0(workingd,"/data/ldsc_output/whole_brain")
setwd(temporarywd)


load("whole_brain.RData")
ldscoutput<-LDSCoutput_wholebrain
whole_brain_corr<-reshape2::melt(ldscoutput$S_Stand)
whole_brain_corr$value_corrected<-ifelse(whole_brain_corr$value>1,1,whole_brain_corr$value)

  heatmap<-
    ggplot(data=whole_brain_corr, aes(Var1,Var2,fill=value_corrected))+
    geom_tile(color = "white")+
    theme_minimal()+ 
    theme_bw()+
    theme(axis.text.x = element_blank())+
    theme(axis.text.y = element_blank())+
    theme(axis.title.x=element_blank(),
          axis.title.y = element_blank(),
          panel.grid.major = element_blank(),
          panel.border = element_blank(),
          axis.ticks = element_blank(),
          legend.justification = c(1,0),
          #legend.position = c(0.45,0.7),
          legend.direction = "horizontal")+  
    scale_fill_gradient2(low="royalblue1",high="palevioletred2",mid ="white",
                         midpoint=0,limit=c(-1,1),na.value="white",
                         name="Genetic\ncorrelation")+
    guides(fill=guide_colorbar(barwidth = 7,barheight = 1,title.position = "top",title.hjust = 0.5))+
    coord_fixed()
  heatmap
```


The vertical line through the diagnonal indicates the genetic correlation of a brain area with itself, and the vertical lines that run parallel to it indicate the strong genetic correlation with their contralateral counterpart.

<br>
<br>

### Code for heatmaps in Supplementary Materials 

```{r heatmaps_for_manuscript, eval=F}
### whole brain
tiff("heatmap_whole_brain.tiff", width = 7, height = 6, units = 'in', res=1000)
heatmap
dev.off()


### central executive
tiff("heatmap_central_exec.tiff", width = 6, height = 6, units = 'in', res=1000)
cor_central_exec_heatmap
dev.off()

### cingulo
tiff("heatmap_cingulo.tiff", width = 6, height = 6, units = 'in', res=1000)
cor_cingulo_heatmap
dev.off()

### default mode
tiff("heatmap_default.tiff", width = 7, height = 7, units = 'in', res=1000)
cor_default_mode_heatmap
dev.off()

### hippocampal
tiff("heatmap_hippocampal.tiff", width = 6, height = 6, units = 'in', res=1000)
cor_hippocampal_heatmap
dev.off()

### Multiple demand
tiff("heatmap_multiple.tiff", width = 6, height = 6, units = 'in', res=1000)
cor_multiple_heatmap
dev.off()


### P-FIT
tiff("heatmap_p_fit.tiff", width = 10, height = 10, units = 'in', res=1000)
cor_p_fit_heatmap
dev.off()

### Salience
tiff("heatmap_salience.tiff", width = 6, height = 6, units = 'in', res=1000)
cor_salience_heatmap
dev.off()

### Sensorimotor
tiff("heatmap_sensorimotor.tiff", width = 6, height = 6, units = 'in', res=1000)
cor_sensori_heatmap
dev.off()

### temporo
tiff("heatmap_temporo.tiff", width = 10, height = 10, units = 'in', res=1000)
cor_temporo_heatmap
dev.off()
```


***

<br>
<br>

# Sanity checks

#### Display genetic correlations
```{r plot_var_corr,message=F,warning=F}
library(plyr)
library(TeachingDemos)
source("https://raw.githubusercontent.com/talgalili/R-code-snippets/master/boxplot.with.outlier.label.r")

# load whole brain data
workingd<-getwd()
temporarywd<-paste0(workingd,"/data/ldsc_output/whole_brain")
setwd(temporarywd)

load("whole_brain.RData")
ldscoutput<-LDSCoutput_wholebrain
  dimnames(ldscoutput$S_Stand)[[1]]<-dimnames(ldscoutput$S)[[2]]
  dimnames(ldscoutput$S_Stand)[[2]]<-dimnames(ldscoutput$S)[[2]]

  lower_triangle<-get_lower_tri(ldscoutput$S_Stand)
  diag(lower_triangle)<-NA
  lower_triangle<-reshape2::melt(lower_triangle)
  
lower_triangle$jointname<-paste(lower_triangle$Var1,lower_triangle$Var2)


boxplot.with.outlier.label(lower_triangle$value,NA,xaxt="n",yaxt="n",ylab="LDSC coefficient",ylim=c(-0.8,1.2),main="Variation in LDSC coefficients\namong 83 traits",spread_text = F)
axis(side=2,las=2)

# display the lower correlations
#lower_triangle$jointname<-ifelse(lower_triangle$value<=-0.25,lower_triangle$jointname,NA)

display<-lower_triangle[which(!is.na(lower_triangle$value)),c("Var1","Var2","value")]
display$value<-round(display$value, digits=2)
names(display)<-c("Volume 1","Volume 2","Genetic correlation")


library(knitr)
#kable()
library(DT)
datatable(display, rownames=FALSE, filter="top",options= list(pageLength=5,scrollX=T))
 # ggplot(lower_triangle, aes(y = lower_triangle$value,x=1)) +
#    geom_boxplot() +
#    geom_text(aes(label = lower_triangle$jointname), na.rm = TRUE, nudge_y = 0.05)
```

***

#### Plot LDSC intercepts

```{r plot_intercepts_code,eval=T,echo=T,message=F,warning=F}
### plot intercepts of genetic correlation matrices
rm(list=ls())
## this script is displaying the heatmaps of the networks

# load dependencies
library(stringr)
library(reshape2)
library(ggplot2)

# load data and name it according to network
workingd<-getwd()
temporarywd<-paste0(workingd,"/data/ldsc_output/")
setwd(temporarywd)

networks<-list.files(pattern=".RData")
network_names<-str_replace(networks,pattern=".RData",replacement = "")

for(i in 1:length(networks)){
load(networks[i])
  name<-network_names[i]
  assign(name,LDSCoutput)
}

## name all columns in $I after brain regions and round $I
for(i in network_names){
  output<-get(i)
  dimnames(output$I)[[1]]<-dimnames(output$S)[[2]]
  dimnames(output$I)[[2]]<-dimnames(output$S)[[2]]
  name<-i
  assign(name,output)
  output$I<-round(output$I,digits = 2)
  name_cor<-paste0("intercepts_",i)
  assign(name_cor,output$I)
}

# count number of correlation matrices
#length(ls(pattern="intercepts_"))

# create vector containing names of the correlation matrices
matrices<-ls(pattern="intercepts_")

## plot the rounded correlation matrices

# get lower triangle of matrix
for(i in matrices){
  cormatrix<- get(i)
  get_lower_tri<-function(cormatrix){
    cormatrix[upper.tri(cormatrix)] <- NA
    return(cormatrix)
  }
  lower_triangle<-get_lower_tri(cormatrix)
  lower_triangle<-reshape2::melt(lower_triangle)
  lower_triangle$corrected<-with(lower_triangle, ifelse(value>=1.00,1.00,value))

  #print(i)
  #print(lower_triangle)
  
  heatmap<-
    ggplot(data=lower_triangle, aes(Var1,Var2,fill=corrected))+
    geom_tile(color = "white")+
    theme_minimal()+ 
    theme_bw()+
    theme(axis.text.x = element_text(angle = 45, vjust = 1, 
                                     size = 8, hjust = 1))+
    theme(axis.text.x = element_text(angle = 45, vjust = 1, 
                                     size = 8, hjust = 1))+
    theme(axis.title.x=element_blank(),
          axis.title.y = element_blank(),
          panel.grid.major = element_blank(),
          panel.border = element_blank(),
          axis.ticks = element_blank(),
          legend.justification = c(1,0),
          #legend.position = c(0.45,0.7),
          legend.direction = "horizontal")+  
    scale_fill_gradient2(low="gold4",high="darkorange3",mid ="gold",
                         midpoint=0,limit=c(-1,1),na.value="white",
                         name="LDSC intercepts")+
    guides(fill=guide_colorbar(barwidth = 7,barheight = 1,title.position = "top",title.hjust = 0.5))+
    coord_fixed()
  
  heatmap<-heatmap+
    geom_text(aes(Var1,Var2,label=value),color="black",size=2)
  
  name<-paste0(i,"_heatmap")
  assign(name,heatmap)
}

```
```{r intercepts1,echo=FALSE, out.width = '80%',warning=F,message=F,fig.cap=c("Central Executive","Cingulo-opercular")}
library(cowplot)
intercepts_central_exec_heatmap
intercepts_cingulo_heatmap
```

```{r intercepts2,echo=FALSE, out.width = '100%',warning=F,message=F,fig.cap="Default Mode"}
intercepts_default_mode_heatmap
```

#### LDSC intercepts across the whole brain
```{r whole_brain_intercepts,eval=T,echo=F}
workingd<-getwd()
temporarywd<-paste0(workingd,"/data/ldsc_output/whole_brain")
setwd(temporarywd)

load("whole_brain.RData")
ldscoutput<-LDSCoutput_wholebrain
whole_brain_inter<-reshape2::melt(ldscoutput$I)
whole_brain_inter$value_corrected<-ifelse(whole_brain_inter$value>1,1,whole_brain_inter$value)

  heatmap<-
    ggplot(data=whole_brain_inter, aes(Var1,Var2,fill=value_corrected))+
    geom_tile(color = "white")+
    theme_minimal()+ 
    theme_bw()+
    theme(axis.text.x = element_blank())+
    theme(axis.text.y = element_blank())+
    theme(axis.title.x=element_blank(),
          axis.title.y = element_blank(),
          panel.grid.major = element_blank(),
          panel.border = element_blank(),
          axis.ticks = element_blank(),
          legend.justification = c(1,0),
          #legend.position = c(0.45,0.7),
          legend.direction = "horizontal")+  
    scale_fill_gradient2(low="gold4",high="darkorange3",mid ="gold",
                         midpoint=0,limit=c(-1,1),na.value="white",
                         name="LDSC intercepts")+
    guides(fill=guide_colorbar(barwidth = 7,barheight = 1,title.position = "top",title.hjust = 0.5))+
    coord_fixed()
  heatmap
```

***

#### Do the LDSC intercepts correlate with the phenotypic correlations between brain areas?
The intercept depend on the phenotypic correlation between traits which can be negative or positive. Thus, we plot the intercepts as a function of the pairwise phenotypic correlation between traits. Based on LDSC theory it should look like a straight line + small variation due to sampling error.

```{r intercepts_phenotypic_cor}
# load intercepts
workingd<-getwd()
temporarywd<-paste0(workingd,"/data/ldsc_output/whole_brain/")
setwd(temporarywd)

load("whole_brain.RData")
ldscoutput<-LDSCoutput_wholebrain
# name rows and columns
dimnames(ldscoutput$I)[[1]]<-dimnames(ldscoutput$S)[[2]]
dimnames(ldscoutput$I)[[2]]<-dimnames(ldscoutput$S)[[2]]

# keep lower matrix
cormatrix<-ldscoutput$I
get_lower_tri<-function(cormatrix){
  cormatrix[upper.tri(cormatrix)] <- NA
  return(cormatrix)
}
intercepts<-get_lower_tri(cormatrix)
diag(intercepts)<-NA
intercepts<-reshape2::melt(intercepts)
names(intercepts)<-c("region1","region2","intercepts")

# load phenotypic correlations
temporarywd<-paste0(workingd,"/data/phenotypic/")
setwd(temporarywd)

load("residualised_Zhao_corr.RData")

phenotypic_cor<-cor_matrix

# sort same as intercept matrix
names<-dimnames(ldscoutput$S)[[2]]
phenotypic_cor<-phenotypic_cor[names,names]

# get lower triangle
phenotypic_cor<-get_lower_tri(phenotypic_cor)
phenotypic_cor<-reshape2::melt(phenotypic_cor)
names(phenotypic_cor)<-c("region1","region2","phenotypic_cor")

intercept_pheno<-merge(intercepts,phenotypic_cor,by=c("region1","region2"))

plot(intercept_pheno$phenotypic_cor,intercept_pheno$intercepts,xlab="Phenotypic correlations",ylab="LDSC intercepts")
abline(lm(intercept_pheno$intercepts~intercept_pheno$phenotypic_cor,na.action = "na.omit"),col="red")
summary(lm(intercept_pheno$intercepts~intercept_pheno$phenotypic_cor))
# how many intercepts are there? Should be 3404
#sum(!is.na(intercepts$intercepts))
```

***

#### Are all traits heritable?
Distribution of heritability estimates for all 83 brain volumes

```{bash univariate_ldsc, eval=F,echo=T}
#!/bin/bash -l

#SBATCH --output=~/genetic_networks/outputs/ldsc/calculate_heritabilities.txt
#SBATCH --job-name=heritability_brain_volume


# specify time limit 
#SBATCH --time=0-99:00

# memory per job 
#SBATCH --mem=12G

module load devtools/anaconda/2019.3-python2.7.16

~/Software/ldscore/ldsc/ldsc.py \
--rg ~/genetic_networks/data/GWAS_munged/Brain_stem.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_accumbens_area.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_amygdala.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_bankssts.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_caudal_anterior_cingulate.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_caudal_middle_frontal.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_caudate.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_cuneus.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_DC.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_entorhinal.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_frontal_pole.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_fusiform.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_hippocampus.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_inferior_parietal.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_inferior_temporal.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_insula.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_isthmus_cingulate.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_lateral_occipital.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_lateral_orbitofrontal.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_lingual.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_medial_orbitofrontal.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_middle_temporal.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_pallidum.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_paracentral.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_parahippocampal.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_pars_opercularis.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_pars_orbitalis.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_pars_triangularis.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_pericalcarine.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_postcentral.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_posterior_cingulate.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_precentral.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_precuneus.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_putamen.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_rostral_anterior_cingulate.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_rostral_middle_frontal.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_superior_frontal.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_superior_parietal.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_superior_temporal.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_supramarginal.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_thalamus_proper.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Left_transverse_temporal.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_accumbens_area.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_amygdala.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_bankssts.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_caudal_anterior_cingulate.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_caudal_middle_frontal.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_caudate.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_cuneus.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_DC.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_entorhinal.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_frontal_pole.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_fusiform.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_hippocampus.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_inferior_parietal.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_inferior_temporal.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_insula.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_isthmus_cingulate.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_lateral_occipital.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_lateral_orbitofrontal.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_lingual.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_medial_orbitofrontal.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_middle_temporal.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_pallidum.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_paracentral.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_parahippocampal.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_pars_opercularis.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_pars_orbitalis.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_pars_triangularis.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_pericalcarine.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_postcentral.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_posterior_cingulate.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_precentral.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_precuneus.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_putamen.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_rostral_anterior_cingulate.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_rostral_middle_frontal.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_superior_frontal.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_superior_parietal.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_superior_temporal.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_supramarginal.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_thalamus_proper.sumstats.gz,\
~/genetic_networks/data/GWAS_munged/Right_transverse_temporal.sumstats.gz \
--ref-ld-chr ~/Software/ldscore/eur_w_ld_chr/ \
--w-ld-chr ~/Software/ldscore/eur_w_ld_chr/ \
--out ~/genetic_networks/outputs/ldsc/volumes_heritability_ldsc.out

### Aim: run univariate LDSC to get heritability estimates with Bulik-Sullivan python software
### extract files in cd ~/genetic_networks/data/GWAS_munged/ and save in a file
### using ls -d "$PWD"/*.gz > listOfFiles.list

### Author: Anna Furtjes
### Date: 16/05/2020
### submit using: sbatch -p shared ~/genetic_networks/scripts/perform_ldsc/ldsc_heritability_volumes_26052020.sh


## copy over onto local machine
```


```{r display_h2, message=F,eval=T,fig.width=5,warning=F}
workingd<-getwd()
temporarywd<-paste0(workingd,"/data_Smith/heritability_volumes/")
setwd(temporarywd)

library(data.table)
heritability<-fread("heritability_brain_volumes_26052020.txt",header=T,data.table = F)
summary(heritability$h2_obs)

library(plyr)
source("https://raw.githubusercontent.com/talgalili/R-code-snippets/master/boxplot.with.outlier.label.r")
boxplot_h2<-boxplot.with.outlier.label(heritability$h2_obs,heritability$p2,xaxt="n",yaxt="n",ylab="SNP heritability",spread_text = F)
axis(side=2,las=2)
```

#### Display heritability estimates
```{r lowest_heritabilities}
table<-heritability[,c("p2","h2_obs","h2_obs_se")]
names(table)<-c("Brain areas","h2 observed","SE")
table<-table[order(table$`h2 observed`),]
library(DT)
datatable(table, rownames=FALSE, filter="top",options = list(pageLength=5, scrollX=T))
```


*Note.* The frontal pole is only a tiny area which often is difficult to segment. This might have caused low levels of systematic variation across the GWAS resulting in very low SNP heritability captured by LDSC.

***

<br>
<br>

# Data display in manuscript

#### Boxplots displaying SNP-heritabilites and genetic correlations 
```{r raincloud_plot, eval=T,message=F,warning=F}
library(readr)
library(tidyr)
library(ggplot2)
library(Hmisc)
library(plyr)
library(RColorBrewer)
library(reshape2)
library(PupillometryR)
library(cowplot)
# plotted with explanation on : https://micahallen.org/2018/03/15/introducing-raincloud-plots/


#tiff("boxplots.tiff", width = 9, height = 6, units = 'in', res=1000)

##########################################
## Plot SNP-heriability
#########################################
# data to be plotted: heritability estimates
heritability$name<-rep("SNP-heritability",83)
heritability$outlier<-ifelse(heritability$p2=="Brain_stem", "Brain stem",NA)

# plot SNP-heritability
plot_heritability<-
                  ggplot(data=heritability,aes(x=name,y=h2_obs))+
                  geom_flat_violin(position = position_nudge(x = .2, y = 0), alpha = .8, fill="gray85",alpha=0.1) +
  geom_point(aes(y = h2_obs), position = position_jitter(width = .15), size = .5, alpha = 1,colour="gray65")+
  geom_boxplot(width = .2, guides = FALSE, outlier.shape = NA, alpha = 0.5)+ 
  geom_text(aes(label = outlier), na.rm = TRUE, hjust = 1.1, vjust=-1.2, colour="gray65",size=3.5)+
   expand_limits(x=2)+
  theme_bw() +
  theme(text = element_text(size=12),
        axis.text.x = element_blank(),
        axis.text.y = element_text(size=13),
        axis.title.y = element_text(face="bold", colour='black', size=17),
        axis.title.x = element_blank())+
  ylab("SNP-heritability")

##################################################
# Plot genetic correlations
#################################################
temporarywd<-paste0(workingd,"/data/ldsc_output/whole_brain/")
setwd(temporarywd)

load("whole_brain.RData")
ldscoutput<-LDSCoutput_wholebrain
  dimnames(ldscoutput$S_Stand)[[1]]<-dimnames(ldscoutput$S)[[2]]
  dimnames(ldscoutput$S_Stand)[[2]]<-dimnames(ldscoutput$S)[[2]]

  get_lower_tri<-function(cormatrix){
  cormatrix[upper.tri(cormatrix)] <- NA
  return(cormatrix)
  }
  
  lower_triangle<-get_lower_tri(ldscoutput$S_Stand)
  # remove correlations with oneself
  diag(lower_triangle)<-NA
  #melt correlation matrix
  lower_triangle<-reshape2::melt(lower_triangle)
  # remove correlations with itself
  lower_triangle$value<-ifelse(lower_triangle$Var1 == lower_triangle$Var2,NA,lower_triangle$value)

lower_triangle$jointname<-paste(lower_triangle$Var1,lower_triangle$Var2)
lower_triangle$name<-rep("genetic corr",nrow(lower_triangle))


# we only keep lower_triangle values that are not missing
lower_triangle<-lower_triangle[which(!is.na(lower_triangle$value)),]

## plot genetic correlations
plot_genetic_corr<-
                  ggplot(data=lower_triangle,aes(x=name,y=value))+
                  geom_flat_violin(position = position_nudge(x = .2, y = 0), alpha = .8,fill="gray85",alpha=0.6) +
                  geom_point(aes(y = value), position = position_jitter(width = .15), size = .5, alpha = 0.9,colour="gray65") +
                  geom_boxplot(width = .2, guides = FALSE, outlier.shape = NA, alpha = 0.5)+
                  expand_limits(x=2)+
                  theme_bw()+
                  theme(text = element_text(size=12),
                        axis.text.x = element_blank(),
                        axis.text.y = element_text(size=13),
                        axis.title.y = element_text(face="bold", colour='black', size=17),
                        axis.title.x = element_blank())+
                  ylab("Genetic correlations")+
                  ylim(-0.4,1)

#############################################################################################
# make same plot for phenotypic correlations
# code for how we obtain the cor_matrix on the Edinburgh server can be found in this document under "Pre-processing: Phenotypic data"
temporarywd_pheno<-paste0(workingd,"/data/phenotypic/")
setwd(temporarywd_pheno)

load("residualised_Zhao_corr.RData")

lower_triangle_pheno<-get_lower_tri(cor_matrix)
# remove correlations with oneself
diag(lower_triangle_pheno)<-NA
#melt
lower_triangle_pheno<-reshape2::melt(lower_triangle_pheno)
lower_triangle_pheno$value<-ifelse(lower_triangle_pheno$Var1 == lower_triangle_pheno$Var2,NA,lower_triangle_pheno$value)

# we only keep non-missing values
lower_triangle_pheno<-lower_triangle_pheno[which(!is.na(lower_triangle_pheno$value)),]

# add variable to plot against
lower_triangle_pheno$name<-rep("phenotypic corr",nrow(lower_triangle_pheno))

## plot phenotypic correlations
plot_phenotypic_corr<-
                  ggplot(data=lower_triangle_pheno,aes(x=name,y=value))+
                  geom_flat_violin(position = position_nudge(x = .2, y = 0), alpha = .8,fill="gray85",alpha=0.6) +
                  geom_point(aes(y = value), position = position_jitter(width = .15), size = .5, alpha = 0.9,colour="gray65") +
                  geom_boxplot(width = .2, guides = FALSE, outlier.shape = NA, alpha = 0.5)+
                  expand_limits(x=2)+
                  theme_bw()+
                  theme(text = element_text(size=12),
                        axis.text.x = element_blank(),
                        axis.text.y = element_text(size=13),
                        axis.title.y = element_text(face="bold", colour='black', size=17),
                        axis.title.x = element_blank())+
                  ylab("Phenotypic correlations")+
                  ylim(-0.4,1)
#############################################################################################
# display all three plots together
plot_grid(plot_heritability,plot_genetic_corr,plot_phenotypic_corr,ncol=3,nrow=1,labels = c("A","B","C"))

#setwd(workingd)
#dev.off()
```


