---
title: "SGDP_Central_Genotype_QC"
author: "Joni Coleman"
date: "20/04/2020"
output: 
  html_document:
    includes:
      after_body: footer.html 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Notes

This RMarkdown is written as a self-contained codebook for the centralised quality control of the genetic data downloaded from the UK Biobank at the Social, Genetic and Developmental Psychiatry [SGDP] Centre, King's College London [KCL]. 
The output of this centralised quality control produces cleaned datasets for use by multiple analysts across KCL.
The scripts are constructed to be usable for the different projects run from KCL, and so are generalisable to UKBB projects overall.
Developments of this code were used in the R package ukbtools [https://cran.r-project.org/web/packages/ukbtools/index.html] - credit: Ken Hanscombe
This document should not be run as a R script itself, but is rather a single document to list all code.
All code was run on the King's College London / National Institute of Health Research Maudsley Biomedical Research Centre high performance computing cluster "Rosalind", unless otherwise stated.

Scripts have been simplified by removing paths to files and renaming certain files to more widely-understandable names where relevant

## Dependencies

No previous scripts
UKBB genotype and imputed genetic data
PLINK 1.9 (see below)
FlashPCA2 (see below)
GreedyRelated (see below)
bgenix (see below)

## Credit

Contributions to the logic of this code were made by numerous analysts at the SGDP.
The code itself was written by Jonathan Coleman and Sam Choi.
Additional programs are the credit of their authors:
    PLINK[https://www.cog-genomics.org/plink/1.9/]
        Chris Chang, Shaun Purcell
    FlashPCA2[https://github.com/gabraham/flashpca]
        Gad Abraham
    GreedyRelated[https://github.com/choishingwan/GreedyRelated]
        Sam Choi
    bgenix[https://bitbucket.org/gavinband/bgen/wiki/bgenix]
	Gavin Band
    
### Genotype data code

## Wrapper for genotype QC

```{sh Run_Genotyped_Data_QC_JRIC_240119.sh, eval=F}

## Note - text in square brackets below should be replaced with suitable file name (don't put the square brackets around the files!)

Rscript Make_UKBB_Caucasians_By_4_Means_Clustering.R \
[File containing the IDs and the 40 UKBB PCs extracted from ...sqc_v2.txt - e.g. ukb1657_40PCs_ID_from_sqc_v2.txt] \
[k for k-means clustering, e.g. 4] \
[Full path to output root, e.g. to/UKBB_Data_Ancestry. Example output will be to/UKBB_Data_Ancestry.txt, to/UKBB_Data_Ancestry.pdf, to/UKBB_Data_Ancestry_PLINK.txt]

bash Genotyped_Data_QC_JRIC_240119.bash \
-b /mnt/lustre/groups/ukbiobank/ukb18177_glanville/genotyped/ukb18177_glanville_binary_pre_qc.bed  \
-j /mnt/lustre/groups/ukbiobank/ukb18177_glanville/genotyped/ukb18177_glanville_binary_pre_qc.bim \
-f /mnt/lustre/groups/ukbiobank/ukb18177_glanville/genotyped/ukb18177_glanville_binary_pre_qc.fam \
-m 0.01 \
-v 0.02 \
-s 0.02 \
-a 1 \
-q 1 \
-e /scratch/groups/ukbiobank/KCL_Data/Genotypes/kylie_application/ukb1817_het_miss_excl_rows_v3.txt \
-g 1 \
-x /scratch/groups/ukbiobank/KCL_Data/Genotypes/kylie_application/ukb1817_reported_sex_from_sqc_v2.txt \
-r 0.044 \
-l /scratch/groups/ukbiobank/KCL_Data/Genotypes/kylie_application/ukb1817_rel_pairs.txt \
-t 1 \
-u /mnt/lustre/groups/ukbiobank/Edinburgh_Data/usr/anna/PhD/output/geno_qc/participants_of_interest.txt \ # this file will have header: ID Pheno
-h 0.00000001 \
-d 0 \
-p 1 \
-c /scratch/groups/ukbiobank/KCL_Data/Genotypes/kylie_application/ukb1817_Caucasians_4Means_PLINK.txt \
-o /mnt/lustre/groups/ukbiobank/Edinburgh_Data/usr/anna/PhD/output/geno_qc/genetic_qc_UKB

```

## Identify White Western European Individuals

```{r Make_UKBB_Caucasians_By_4_Means_Clustering.R, eval=F}

##Load dependencies
library(data.table)
library(ggplot2)

args<-commandArgs(trailingOnly=TRUE)

if(length(args)==0){
    stop("Please provide an input file containing an ID column and all PCs (full path), K for k-means clustering, and an output file path", call.=FALSE)
}

##Read in PCs, remove NAs, rename
PCs<-fread(args[1], data.table=F)
PCs<-na.omit(PCs)
names(PCs)<-c("ID","PC.1","PC.2","PC.3","PC.4","PC.5","PC.6","PC.7","PC.8","PC.9","PC.10","PC.11","PC.12","PC.13","PC.14","PC.15","PC.16","PC.17","PC.18","PC.19","PC.20","PC.21","PC.22","PC.23","PC.24","PC.25","PC.26","PC.27","PC.28","PC.29","PC.30","PC.31","PC.32","PC.33","PC.34","PC.35","PC.36","PC.37","PC.38","PC.39","PC.40")

##Set seed

set.seed(1204688)

##K means clustering on each PC

K_MEAN<-as.numeric(args[2])

PC1_K<-kmeans(PCs$PC.1, K_MEAN)
PC2_K<-kmeans(PCs$PC.2, K_MEAN)

##Add clusters to PC dataframe
PCs$PC1.Cluster<-PC1_K$cluster
PCs$PC2.Cluster<-PC2_K$cluster
PCs$Clusters<-as.factor(paste(PC1_K$cluster,PC2_K$cluster,sep="."))

##WWE group is the majority

MAX_PC1<-ifelse(match(max(table(PCs$PC1.Cluster, PCs$PC2.Cluster)), table(PCs$PC1.Cluster, PCs$PC2.Cluster)) %% K_MEAN == 0, K_MEAN, match(max(table(PCs$PC1.Cluster, PCs$PC2.Cluster)), table(PCs$PC1.Cluster, PCs$PC2.Cluster)) %% K_MEAN)

MAX_PC2<-ceiling(match(max(table(PCs$PC1.Cluster, PCs$PC2.Cluster)), table(PCs$PC1.Cluster, PCs$PC2.Cluster))/K_MEAN)

##Make lists of WWE IDs
WWE<-as.data.frame(PCs[PCs$PC1.Cluster == MAX_PC1 & PCs$PC2.Cluster == MAX_PC2,1])
names(WWE)<-"ID"

WWE_PLINK<-as.data.frame(cbind(WWE,WWE))
names(WWE_PLINK)<-c("FID","IID")

##Write to file
write.table(WWE, file=paste(args[3],".txt",sep=""), row.names=F, col.names=T, quote = F)
write.table(WWE_PLINK, file=paste(args[3],"_PLINK.txt",sep=""), row.names=F, col.names=T, quote = F)

##Plot
pdf(paste(args[3], ".pdf", sep=""))
with(PCs, print(qplot(PC.1, PC.2, colour=Clusters)))
dev.off()
```

## QC genotyped data

```{bash Genotyped_Data_QC_JRIC_240119.bash, eval=F}

#Check inputs are suitable

while getopts ":b:j:f:m:v:s:a:q:e:g:x:r:l:t:u:h:d:p:c:o:" opt
do
    case $opt in
	b) INPUT_BED="$OPTARG"
	;;
	j) INPUT_BIM="$OPTARG"
	;;
	f) INPUT_FAM="$OPTARG"
	;;
        m) MAF="$OPTARG"
        ;;
        v) GENO="$OPTARG"
        ;;
        s) MIND="$OPTARG"
        ;;
        a) CAUC="$OPTARG"
	;;
	q) QA="$OPTARG"
	;;
	e) QA_FILE="$OPTARG"
	;;
	g) GENDER="$OPTARG"
	;;
	x) SEX_FILE="$OPTARG"
	;;
	r) RELATEDNESS="$OPTARG"
	;;
	l) RELATE_FILE="$OPTARG"
	;;
	t) RELATE_PHENO="$OPTARG"
	;;
	u) RELATE_PHENO_FILE="$OPTARG"
	;;
        h) HWE="$OPTARG"
        ;;
	p) PRUNE="$OPTARG"
	;;
	d) SEXSPLIT="$OPTARG"
	;;
	c) CAUCASIANS_FILE="$OPTARG"
	;;
	o) OUTPUT="$OPTARG"
	;;
        \?) echo "Invalid option -$OPTARG" >&2
        ;;
    esac
done

if [ $# -ne 40 ]
then
    echo -e "\nPlease provide all arguments. If argument not needed:\n -b [Full path to input bed]\n -j [Full path to input bim]\n -f [Full path to input fam]\n -m [MAF]\n -v [GENO]\n -s [MIND]\n -a [Exclude non-Caucasians? 1 = Yes, 0 = No]\n -q [Exclude QA errors? 1 = Yes, 0 = No]\n -e [Full path to UKBB QA Excludes file]\n -g [Exclude gender mismatches? 1 = Yes, 0 = No]\n -x [Full path to UKBB Phenotypic Sex file]\n -r [Relatedness threshold for greedyRelated]\n -l [Full path to Relatedness Pairs file]\n -t [Use phenotype file to retain cases preferentially? 1 = Yes 0 = No]\n -u [Full path to phenotype file for Relatedness pairs]\n -h [HWE for SNPs]\n -d [Create sex split? 1 = Yes, 0 = No]\n -p [Create pruned file? 1 = Yes, 0 = No]\n -c [Full path to Caucasians File]\n -o [Full path to output root]\nArguments provided = $#"
    exit
fi

## MAF

MAF_FLAG=$(echo ${MAF} != 0 | bc -l)
MIND_FLAG=$(echo ${MIND} != 0 | bc -l)
RELATE_FLAG=$(echo ${RELATEDNESS} != 1 | bc -l)

first_plink="plink --bed ${INPUT_BED} --bim ${INPUT_BIM} --fam ${INPUT_FAM} --geno ${GENO} --write-snplist --out ${OUTPUT}_GENO${GENO}"

$first_plink

## Keep {OUTPUT}_GENO${GENO}.fam

plink="plink --bed ${INPUT_BED} --bim ${INPUT_BIM} --fam ${INPUT_FAM} --geno ${GENO} --extract ${OUTPUT}_GENO${GENO}.snplist --write-snplist --make-just-fam"

if [ ${MAF_FLAG} -ne 0 ]
then 
    plink="${plink} --freq --maf ${MAF}"
fi

## Sample Missingness

if [ ${MIND_FLAG} -eq 1 ]
then
    echo -e "\nWARNING: You are applying a missingness filter to samples.\n If your base file is not a whole-genome file, this may produce different sample sets between clumps/chromosomes.\n"

    plink="${plink} --mind ${MIND}"
fi

if [ ${CAUC} -eq 1 ]
then
    ## Keep {OUTPUT}_GENO${GENO}.fam and Caucasians file
    plink="${plink} --keep ${CAUCASIANS_FILE}"
fi

if [ ${CAUC} -eq 0 ]
then
    echo -e "\nWARNING: You are retaining all individuals - this will produce a transethnic dataset \n"
fi

##Exclude relatives greedily 

if [ ${RELATE_FLAG} -eq 0 ]
then
    echo -e "\nWARNING: You are not removing related individuals \n"
fi

if [ ${RELATE_FLAG} -ne 0 ]
then
   if [ ${QA} -eq 1 ]
   then
      ##Get plink .fam for exclusions to date, including QA
      plink_pre_relate="${plink} --remove ${QA_FILE} --out ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}"

      $plink_pre_relate
   fi

   if [ ${QA} -eq 0 ]
   then
      ##Get plink .fam for exclusions to date, including QA
      plink_pre_relate_no_QA="${plink} --out ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}"

      $plink_pre_relate_no_QA
   fi

   ##Limit relatedness file to included participants
   cat <(head -1 ${RELATE_FILE}) <(LANG=C fgrep -wf <(awk '{print $1}' ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}.fam) ${RELATE_FILE}) > ${RELATE_FILE}_reduced

   if [ ${RELATE_PHENO} -eq 0 ]
   then
      greedRelate/greedyRelated \
      -r ${RELATE_FILE}_reduced \
      -t ${RELATEDNESS} \
      -s 1234 \
      > ${OUTPUT}_KCL_UKBB_Relatedness_Exclusions_Greedy_1234_${RELATEDNESS}.txt
   fi
   if [ ${RELATE_PHENO} -ne 0 ]
   then
      greedRelate/greedyRelated \
      -r ${RELATE_FILE}_reduced \
      -p ${RELATE_PHENO_FILE} \
      -t ${RELATEDNESS} \
      -s 1234 \
      > ${OUTPUT}_KCL_UKBB_Relatedness_Exclusions_Greedy_1234_${RELATEDNESS}.txt
   fi      
   cat <(echo -e "FID IID") <(awk '{print $1, $1}' ${OUTPUT}_KCL_UKBB_Relatedness_Exclusions_Greedy_1234_${RELATEDNESS}.txt) > ${OUTPUT}_KCL_UKBB_Relatedness_Exclusions_Greedy_1234_${RELATEDNESS}_FOR_PLINK.txt
fi

## Exclude all UKB exclusions (UKBileve_Affy... + UKBileve_Geno... + Recommended Excl... For details of these exlcusions, see http://biobank.ctsu.ox.ac.uk/showcase/search.cgi?wot=0&srch=quality+&sta0=on&sta1=on&sta2=on&sta3=on&str0=on&str3=on&fit0=on&fit10=on&fit20=on&fit30=on&fvt11=on&fvt21=on&fvt22=on&fvt31=on&fvt41=on&fvt51=on&fvt61=on&fvt101=on)

if [ ${QA} -eq 1 ]
then

    if [ ${RELATE_FLAG} -ne 0 ]
    then
	cat ${QA_FILE} ${OUTPUT}_KCL_UKBB_Relatedness_Exclusions_Greedy_1234_${RELATEDNESS}_FOR_PLINK.txt | sort | uniq > ${OUTPUT}_KCL_UKBB_QA_And_Relatedness_Exclusions_FOR_PLINK.txt
        plink="${plink} --remove ${OUTPUT}_KCL_UKBB_QA_And_Relatedness_Exclusions_FOR_PLINK.txt --out ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE} --hardy --hwe ${HWE}"
    fi

    if [ ${RELATE_FLAG} -eq 0 ]
    then
	plink="${plink} --remove ${QA_FILE} --out ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE} --hardy --hwe ${HWE}"
    fi
fi

if [ ${QA} -eq 0 ]
then
    echo -e "\nWARNING: You are not removing UKBiobank QA exclusions - ARE YOU SURE? \n"
    if [ ${RELATE_FLAG} -ne 0 ]
    then
        plink="${plink} --remove ${OUTPUT}_KCL_UKBB_Relatedness_Exclusions_Greedy_1234_${RELATEDNESS}_FOR_PLINK.txt --out ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE} --hardy --hwe ${HWE}"
    fi

    if [ ${RELATE_FLAG} -eq 0 ]
    then
        plink="${plink} --out ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE} --hardy --hwe ${HWE}"
    fi

fi

##Prune

###JRIC - 020617 - This command previously included Split-X. This makes a new file, which we want to avoid. I have moved this to the start of the analysis so that this is run first time (if needed)

#Run the previous QC first
$plink

if [ $PRUNE -eq 1 ]
then

    awk -f gwas_scripts/highLDregions4bim_b37.awk ${INPUT_BIM} > ${OUTPUT}_High_LD_Regions_To_Exclude.txt

    plink \
    --bed ${INPUT_BED} \
    --bim ${INPUT_BIM} \
    --fam ${INPUT_FAM} \
    --maf 0.05 \
    --hwe 0.001 \
    --geno 0.02 \
    --thin-indiv-count 300 \
    --indep-pairphase 200 100 0.2 \
    --seed 1204688 \
    --exclude ${OUTPUT}_High_LD_Regions_To_Exclude.txt \
    --extract ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}.snplist \
    --keep ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}.fam \
    --out ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}_LD1_ALL

    plink \
    --bed ${INPUT_BED} \
    --bim ${INPUT_BIM} \
    --fam ${INPUT_FAM} \
    --autosome \
    --maf 0.05 \
    --hwe 0.001 \
    --geno 0.02 \
    --thin-indiv-count 300 \
    --indep-pairphase 200 100 0.2 \
    --seed 1204688 \
    --exclude ${OUTPUT}_High_LD_Regions_To_Exclude.txt \
    --extract ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}.snplist \
    --keep ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}.fam \
    --out ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}_LD1_NO_X

    if [ ${GENDER} -eq 1 ]
    then 
        sex_plink="plink --bed ${INPUT_BED} --bim ${INPUT_BIM} --fam ${INPUT_FAM} --check-sex --keep ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}.fam --extract ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}_LD1_ALL.prune.in --out ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}_PRUNE_SEX"
	$sex_plink
	##Compare to UKBB_Sex - no males have F <0.9, some females with high Fs, but this is possible, so no exclusions for now
	cat <(echo -e "FID SEX IID PEDSEX SNPSEX STATUS F") <(join -1 1 -2 1 <(sort -k1,1 ${SEX_FILE}) <(tail -n+2 ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}_PRUNE_SEX.sexcheck | sort -k1,1)) > ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}_SEXCHECK_WITH_UKBB_SEX.sexcheck
	##Deviations - exclude phenotypic males with F < 0.9, phenotypic females with F > 0.6
	#Reduce file read by 1
	cat <(awk '( NR >1 && $2 == "F" && $7 < 0.5)||(NR >1 && $2 == "M" && $7 > 0.9){ print $1, $3;}' ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}_SEXCHECK_WITH_UKBB_SEX.sexcheck) > ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}_SEX${GENDER}.samplelist	
    fi
    
    if [ $GENDER -eq 0 ]
    then
        echo -e "\nWARNING: You are not removing mismatched genders - ARE YOU SURE? \n"
    fi
fi

if [ $PRUNE -ne 1 ]
then
    if [ $GENDER -eq 1 ]
    then
        echo -e "\nWARNING: You are not performing pruning before checking mismatched genders - ARE YOU SURE? \n"
        plink \
        --bed ${INPUT_BED} \
        --bim ${INPUT_BIM} \
        --fam ${INPUT_FAM} \
	--extract ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}.snplist \
	--keep ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}.fam \
	--check-sex \
	--out ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}_PRUNE_SEX
	
	cat <(echo -e "FID SEX IID PEDSEX SNPSEX STATUS F") <(join -1 1 -2 1 <(sort -k1,1 ${SEX_FILE}) <(tail -n+2 ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}_PRUNE_SEX.sexcheck)) > ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}_SEXCHECK_WITH_UKBB_SEX.sexcheck

        cat <(awk '( NR >1 && $2 == 0 && $7 < 0.6)||(NR >1 && $2 == 1 && $7 > 0.85){ print $1, $3;}' ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}_SEXCHECK_WITH_UKBB_SEX.sexcheck) > ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}_SEX${GENDER}.samplelist		
    fi

    if [ $GENDER -eq 0 ]
    then
        echo -e "\nWARNING: You are not removing mismatched genders - ARE YOU SURE? \n"
    fi
fi

##Sex split

if [ $SEXSPLIT -eq 1 ]
then
    if [ $GENDER -eq 1 ]
    then
        plink \
        --bed ${INPUT_BED} \
	--bim ${INPUT_BIM} \
	--fam ${INPUT_FAM} \
        --extract ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}.snplist \
        --keep ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}_SEX${GENDER}.samplelist \
        --make-just-fam \
	--write-snplist \
        --out ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}_SEX${GENDER}

        plink \
        --bed ${INPUT_BED} \
	--bim ${INPUT_BIM} \
	--fam ${INPUT_FAM} \
        --extract ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}.snplist \
        --keep ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}_SEX${GENDER}.samplelist \
        --filter-males \
        --make-just-fam \
        --out ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}_SEX${GENDER}_MALES
        
        plink \
        --bed ${INPUT_BED} \
	--bim ${INPUT_BIM} \
	--fam ${INPUT_FAM} \
        --extract ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}.snplist \
        --keep ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}_SEX${GENDER}.samplelist \
        --filter-females \
        --make-just-fam \
        --out ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}_SEX${GENDER}_FEMALES
    fi
    if  [ $GENDER -eq 0 ]
    then
        plink \
        --bed ${INPUT_BED} \
	--bim ${INPUT_BIM} \
	--fam ${INPUT_FAM} \
        --extract ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}.snplist \
        --keep ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}.fam \
        --make-just-fam \
	--write-snplist \
        --out ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}_SEX${GENDER}

        plink \
        --bed ${INPUT_BED} \
	--bim ${INPUT_BIM} \
	--fam ${INPUT_FAM} \
        --extract ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}.snplist \
        --keep ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}.fam \
        --filter-males \
        --make-just-fam \
        --out ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}_SEX${GENDER}_MALES

        plink \
        --bed ${INPUT_BED} \
	--bim ${INPUT_BIM} \
	--fam ${INPUT_FAM} \
        --extract ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}.snplist \
        --keep ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}.fam \
        --filter-females \
        --make-just-fam \
        --out ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}_SEX${GENDER}_FEMALES
    fi
fi
if [ $SEXSPLIT -eq 0 ]
then
    if [ $GENDER -eq 1 ]
    then
        plink \
        --bed ${INPUT_BED} \
	--bim ${INPUT_BIM} \
	--fam ${INPUT_FAM} \
        --extract ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}.snplist \
        --keep ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}_SEX${GENDER}.samplelist \
        --make-just-fam \
	--write-snplist \
        --out ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}_SEX${GENDER}
    fi
    if  [ $GENDER -eq 0 ]
    then
        plink \
        --bed ${INPUT_BED} \
	--bim ${INPUT_BIM} \
	--fam ${INPUT_FAM} \
        --extract ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}.snplist \
        --keep ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}.fam \
        --make-just-fam \
	--write-snplist \
        --out ${OUTPUT}_MAF${MAF}_GENO${GENO}_MIND${MIND}_CAUC${CAUC}_UKBQC${QA}_UNREL${RELATEDNESS}_HWE${HWE}_SEX${GENDER}
    fi
fi

```

### Imputed data code 

UKBB data as received is vast and larger than most users will use for common-variant analyses. Thin the UKBB data to MAF >= 0.01 and INFO >= 0.4.

```{sh Extract_MAF1_INFO4_From_Imputed_Data_Updated.sh, eval=F}
awk '$6 >= 0.01 && $8 >= 0.4 {print $2}' ukb_mfi_chr1_v3.txt > ukb_mfi_chr1_v3_MAF1_INFO4.txt
awk '$6 >= 0.01 && $8 >= 0.4 {print $2}' ukb_mfi_chr2_v3.txt > ukb_mfi_chr2_v3_MAF1_INFO4.txt
awk '$6 >= 0.01 && $8 >= 0.4 {print $2}' ukb_mfi_chr3_v3.txt > ukb_mfi_chr3_v3_MAF1_INFO4.txt
awk '$6 >= 0.01 && $8 >= 0.4 {print $2}' ukb_mfi_chr4_v3.txt > ukb_mfi_chr4_v3_MAF1_INFO4.txt
awk '$6 >= 0.01 && $8 >= 0.4 {print $2}' ukb_mfi_chr5_v3.txt > ukb_mfi_chr5_v3_MAF1_INFO4.txt
awk '$6 >= 0.01 && $8 >= 0.4 {print $2}' ukb_mfi_chr6_v3.txt > ukb_mfi_chr6_v3_MAF1_INFO4.txt
awk '$6 >= 0.01 && $8 >= 0.4 {print $2}' ukb_mfi_chr7_v3.txt > ukb_mfi_chr7_v3_MAF1_INFO4.txt
awk '$6 >= 0.01 && $8 >= 0.4 {print $2}' ukb_mfi_chr8_v3.txt > ukb_mfi_chr8_v3_MAF1_INFO4.txt
awk '$6 >= 0.01 && $8 >= 0.4 {print $2}' ukb_mfi_chr9_v3.txt > ukb_mfi_chr9_v3_MAF1_INFO4.txt
awk '$6 >= 0.01 && $8 >= 0.4 {print $2}' ukb_mfi_chr10_v3.txt > ukb_mfi_chr10_v3_MAF1_INFO4.txt
awk '$6 >= 0.01 && $8 >= 0.4 {print $2}' ukb_mfi_chr11_v3.txt > ukb_mfi_chr11_v3_MAF1_INFO4.txt
awk '$6 >= 0.01 && $8 >= 0.4 {print $2}' ukb_mfi_chr12_v3.txt > ukb_mfi_chr12_v3_MAF1_INFO4.txt
awk '$6 >= 0.01 && $8 >= 0.4 {print $2}' ukb_mfi_chr13_v3.txt > ukb_mfi_chr13_v3_MAF1_INFO4.txt
awk '$6 >= 0.01 && $8 >= 0.4 {print $2}' ukb_mfi_chr14_v3.txt > ukb_mfi_chr14_v3_MAF1_INFO4.txt
awk '$6 >= 0.01 && $8 >= 0.4 {print $2}' ukb_mfi_chr15_v3.txt > ukb_mfi_chr15_v3_MAF1_INFO4.txt
awk '$6 >= 0.01 && $8 >= 0.4 {print $2}' ukb_mfi_chr16_v3.txt > ukb_mfi_chr16_v3_MAF1_INFO4.txt
awk '$6 >= 0.01 && $8 >= 0.4 {print $2}' ukb_mfi_chr17_v3.txt > ukb_mfi_chr17_v3_MAF1_INFO4.txt
awk '$6 >= 0.01 && $8 >= 0.4 {print $2}' ukb_mfi_chr18_v3.txt > ukb_mfi_chr18_v3_MAF1_INFO4.txt
awk '$6 >= 0.01 && $8 >= 0.4 {print $2}' ukb_mfi_chr19_v3.txt > ukb_mfi_chr19_v3_MAF1_INFO4.txt
awk '$6 >= 0.01 && $8 >= 0.4 {print $2}' ukb_mfi_chr20_v3.txt > ukb_mfi_chr20_v3_MAF1_INFO4.txt
awk '$6 >= 0.01 && $8 >= 0.4 {print $2}' ukb_mfi_chr21_v3.txt > ukb_mfi_chr21_v3_MAF1_INFO4.txt
awk '$6 >= 0.01 && $8 >= 0.4 {print $2}' ukb_mfi_chr22_v3.txt > ukb_mfi_chr22_v3_MAF1_INFO4.txt
awk '$6 >= 0.01 && $8 >= 0.4 {print $2}' ukb_mfi_chrX_v3.txt > ukb_mfi_chrX_v3_MAF1_INFO4.txt
awk '$6 >= 0.01 && $8 >= 0.4 {print $2}' ukb_mfi_chrXY_v3.txt > ukb_mfi_chrXY_v3_MAF1_INFO4.txt

for i in {22..1} X XY
do
    bgen_tools/bin/bgenix -g ukb_imp_chr${i}_v3.bgen -i ukb_imp_chr${i}_v3.bgen.bgi -incl-rsids ukb_mfi_chr${i}_v3_MAF1_INFO4.txt > ukb_imp_chr${i}_v3_MAF1_INFO4.bgen

    bgen_tools/bin/bgenix -g ukb_imp_chr${i}_v3_MAF1_INFO4.bgen -index
done

```
